A Tale of Two Transects

### 2020-01-22 writing/transect_paper.Rmd
add table of read and ASV counts through pipeline

datasets <- read_csv(here("config/datasets.csv"), col_types = "cccicccccicc")
regions <- read_csv(here("config/regions.csv"), col_types = "cccciiiiic")

@@

region_table <- reads_table %>%
  dplyr::group_by(seq_run, region) %>%
  dplyr::summarize(
    ASVs = dplyr::n_distinct(seq),
    length_min = min(length),
    length_q1 = reldist::wtd.quantile(length, 0.25, weight = reads),
    length_med = reldist::wtd.quantile(length, 0.5, weight = reads),
    length_mean = weighted.mean(length, w = reads),
    length_q3 = reldist::wtd.quantile(length, 0.75, weight = reads),
    length_max = max(length),
    reads = sum(reads)
  )
```

@@

We attempted to denoise both long and short amplicons using `DADA2` version `r packageVersion("dada2")` [@callahan2016; @callahan2019] according to the steps outlined in  the supplementary information in @callahan2019
However, only `r region_table$ASVs[region_table$region=="long"]` amplicon sequence variants (ASVs), were obtained.
We believe that this was due to `DADA2`'s requirement that the seed sequence of each ASV be represented by at least two error-free reads.
If sequencing errors are uniformly distributed, then the probability that a given read will be error-free is $(1-\epsilon)^L$, where $\epsilon$ is the sequencing error rate and $L$ is the read length in base pairs.
Then the number of reads of a given sequence that would be required to obtain two error-free reads in expectation is $2/(1-\epsilon)^L$.
For the combination of long reads (median *X* bp after trimming) and moderate error rate (*X*% mean ccs quality score) in this study, the expected number of reads to achieve two error-free reads is `r round(2/(0.995)^1550)`.
In comparison, the equivalent requirement for the short amplicon (median *X* bp after trimming, mean ccs quality score *Y*%) is only `r round(2/(0.999)^350)`
Given the high diversity relative to sequencing depth in this study (*X* ASVs from PacBio short amplicons, *Y* total long amplicon reads), this requirement was not met for the long amplicons except by the most abundant sequences.
We therefore developed a new workflow to assemble ASVs from the long amplicons, as follows:

@@

Taxonomic annotations of the Ribosomal Data Project's LSU fungal training set (RDP) version 11.5 [@cole2014] and Warcup ITS training set [@wang2007] were mapped to the taxonomic classification system used in the Unite database version 8 [@nilsson2019a].
In particular, the classification for fungi was according to @tedersoo2018, and for non-fungal eukaryotes was according to the proposed system of @tedersoo2017c.
Although the latter system is not formally published, it seems to be in use for non-fungal eukaryotes in the Unite database.
Additionally, it is a system with only purportedly monophyletic taxa and a uniform set of taxon ranks, which make it more appropriate for sequence-based taxonomic assignment algorithms than more accepted classification systems such as @adl2019.
FASTA format files of the re-annotated RDP and Warcup training sets are available at (*somewhere*).

@@

An attempt was made to refine the long amplicon alignment using Pasta [@mirarab2014a], but the resulting maximum likelihood (ML) trees did not improve on the ML achieved using the original alignment from DECIPHER.
Since ML is the optimization criterion used by Pasta, the original DECIPHER alignment was retained.

@@

Long (short) amplicon sequencing with PacBio yielded *X* (*Y*) circular consensus reads, of which *A* (*B*) were successfully demultiplexed.
The 5.8S region was detected in *C* (*D*) reads.
Ion Torrent sequencing yielded *E* demultiplexed reads, with the 5.8S region detected in *F*.

```{r read-counts}

@@

```{r pb-vs-it, fig.cap="(ref:readcap)", fig.height=8}
loadd(big_seq_table_ITS2, cache = cache)
asv_table <- big_seq_table_ITS2 %>%
  tibble::as_tibble(rownames = "filename") %>%
  tidyr::gather(key = "seq", value = "reads", -1) %>%
  dplyr::filter(reads >= 1) %>%
  tidyr::extract(col = "filename",
                 into = c("seq.run", "plate", "well", "dir", "region"),
                 regex = "([a-z]+_\\d+)_(\\d+)_([A-H]1?[0-9])([fr]?)_([:alnum:]+).+") %>%
  dplyr::group_by(seq.run, seq) %>%
  dplyr::summarize(reads = sum(reads)) %>%
  tidyr::spread(key = seq.run, value = reads, fill = 0)

otu <- read_tsv(
  here("data/clusters/ITS2.table"),
  col_types = cols(
    .default = col_integer(),
    `#OTU ID` = col_character()
  )
) %>%
  column_to_rownames("#OTU ID") %>%
  as.matrix() %>%
  t() %>%
  as_tibble(rownames = "sample") %>%
  tidyr::extract(
    col = "sample",
    into = c("seq_run", "plate", "well"),
    regex = "([a-z]{2}_\\d{3})_(\\d{3})([A-H]1?\\d)"
  ) %>%
  tidyr::gather(key = "seq", value = "reads", -(1:3)) %>%
  dplyr::filter(reads >= 1) %>%
  dplyr::group_by(seq_run, seq) %>%
  dplyr::summarize(reads = sum(reads)) %>%
  tidyr::spread(key = seq_run, value = reads, fill = 0)

big_table <- bind_rows(
  mutate(asv_table, type = "ASV"),
  mutate(otu, type = "OTU")
) %>%
  select(-seq)

multi_table <- crossing(
  x_var = factor(datasets$seq_run, ordered = TRUE),
  y_var = factor(datasets$seq_run, ordered = TRUE),
  type = c("ASV", "OTU")
) %>%
  filter(x_var < y_var) %>%
  mutate_all(as.character) %>%
  group_by(x_var, y_var, type) %>%
  group_modify(
    function(d, g) {
      big_table %>%
        filter(type == g$type) %>%
        select(x = !!g$x_var, y = !!g$y_var) %>%
        filter(x > 0 | y > 0)
    }
  )

@@

```{r variofit}
variog_data %>%
  mutate_at("variofitST", lapply, purrr::compose(as.data.frame, as.list, gstat::extractPar)) %>%
  unnest(variofitST) %>%
  arrange(metric, amplicon, guild) %>%
  select(Guild = guild, Amplicon = amplicon, Metric = metric,
         Range = range, Anisotropy = anis) %>%
  mutate(Range = 3*Range,
         "R/A" = Range/Anisotropy,
         "R/A" = ifelse(`R/A` > 100, NA, `R/A`)) %>%
  mutate_at(c("Range", "Anisotropy", "R/A"), formatC, digits = 1,
            format = "f") %>%
  mutate_at("Range", paste, "m") %>%
  mutate_at("Anisotropy", paste, "m/a") %>%
  mutate_at("R/A", paste, "a") %>%
  mutate_at("R/A", str_replace, "NA a", "-") %>%
  mutate_at("Amplicon", fct_relabel, str_replace, " amplicon", "") %>%
  kable(booktabs = TRUE, caption = "(ref:variofit)")
```

@@

```{r phylotax, fig.height = 2, fig.width = 3, fig.cap = "(ref:phylotax)"}
extree <-
  tidytree::as.treedata(ape::read.tree(text = "(A:1,((B:1,C:1):1,((E:1,F:1):1,D:1):1):1);")) %>%
  tidytree::full_join(
    tibble(label = LETTERS[1:6],
           tax1 = c("unk", "Tax1", "Tax2", "Tax2", "unk", "unk"),
           tax2 = c("unk", "Tax2", "Tax2", "Tax1", "unk", "Tax1")),
    by = "label"
  )
ggtree::ggtree(extree) +
  ggtree::geom_tiplab(x = 4, align = 2) +
  xlim(0, 6) +
  ggtree::geom_tiplab(
    aes(color = tax1, label = tax1),
    x = 4.7,
    hjust = 0.5
  ) +
  ggtree::geom_tiplab(
    aes(color = tax2, label = tax2),
    x = 5.7,
    hjust = 0.5
  ) +
  ggtree::geom_tiplab(x = 5.2, label = "/", hjust = 0.5) +
  ggtree::geom_nodelab(aes(label = node - 6), geom = "label") +
  ggtree::geom_hilight(node = 10, fill = "steelblue", extendto = 4) +
  ggtree::geom_hilight(node = 9, fill = "tomato", extendto = 4) +
  scale_color_manual(values = c(Tax1 = "steelblue", Tax2 = "tomato", unk = "grey50")) +
  scale_y_reverse()
```

>>>>

datasets <- read_csv(here("config/datasets.csv"), col_types = "ccccicccccicc")
regions <- read_csv(here("config/regions.csv"), col_types = "cccciiiiic")

@@

region_table <- reads_table %>%
  dplyr::group_by(seq_run, region) %>%
  dplyr::summarize(
    ASVs = dplyr::n_distinct(seq),
    length_min = min(length),
    length_q1 = reldist::wtd.quantile(length, 0.25, weight = reads),
    length_med = reldist::wtd.quantile(length, 0.5, weight = reads),
    length_mean = weighted.mean(length, w = reads),
    length_q3 = reldist::wtd.quantile(length, 0.75, weight = reads),
    length_max = max(length),
    reads = sum(reads)
  )
loadd(big_seq_table_ITS2, cache = cache)
asv_table <- big_seq_table_ITS2 %>%
  tibble::as_tibble(rownames = "filename") %>%
  tidyr::gather(key = "seq", value = "reads", -1) %>%
  dplyr::filter(reads >= 1) %>%
  tidyr::extract(col = "filename",
                 into = c("seq.run", "plate", "well", "dir", "region"),
                 regex = "([a-z]+_\\d+)_(\\d+)_([A-H]1?[0-9])([fr]?)_([:alnum:]+).+") %>%
  dplyr::group_by(seq.run, seq) %>%
  dplyr::summarize(reads = sum(reads)) %>%
  tidyr::spread(key = seq.run, value = reads, fill = 0)

otu <- read_tsv(
  here("data/clusters/ITS2.table"),
  col_types = cols(
    .default = col_integer(),
    `#OTU ID` = col_character()
  )
) %>%
  column_to_rownames("#OTU ID") %>%
  as.matrix() %>%
  t() %>%
  as_tibble(rownames = "sample") %>%
  tidyr::extract(
    col = "sample",
    into = c("seq_run", "plate", "well"),
    regex = "([a-z]{2}_\\d{3})_(\\d{3})([A-H]1?\\d)"
  ) %>%
  tidyr::gather(key = "seq", value = "reads", -(1:3)) %>%
  dplyr::filter(reads >= 1) %>%
  dplyr::group_by(seq_run, seq) %>%
  dplyr::summarize(reads = sum(reads)) %>%
  tidyr::spread(key = seq_run, value = reads, fill = 0)
```

@@

```{r length-qual}

parse_qstat <- function(d) {
  mutate_at(d, "file", basename) %>%
  tidyr::extract(col = "file", into = c("seq_run", "plate", "well"),
                 regex = "([pi][bs]_\\d{3})_(\\d{3})-?([A-H]1?\\d)?[rf]?[.].*")
}

demuxlength <- readd(qstats_length, cache = cache) %>%
  parse_qstat() %>%
  filter(is.na(step), !is.na(well)) %>%
  group_by(seq_run) %>%
  summarize(length = reldist::wtd.quantile(length, weight = nreads)) %>%
  deframe()

demuxqual <- readd(qstats_erate, cache = cache) %>%
  parse_qstat() %>%
  filter(is.na(step), !is.na(well)) %>%
  group_by(seq_run) %>%
  summarize(qual = round(weighted.mean(erate, w = nreads, na.rm = TRUE), 4)) %>%
  deframe()

readcounts <- 
  readd(qstats_n, cache = cache) %>%
  parse_qstat()

rawcounts <- readcounts %>%
  filter(is.na(step), is.na(well)) %>%
  group_by(seq_run) %>%
  summarize(nreads =  prettyNum(sum(nreads), big.mark = " ")) %>%
  deframe()

demuxcounts <- readcounts %>%
  filter(is.na(step), !is.na(well)) %>%
  group_by(seq_run) %>%
  summarize(nreads = prettyNum(sum(nreads), big.mark = " ")) %>%
  deframe()

regioncounts <- readcounts %>%
  filter(step == "lsux") %>%
  group_by(seq_run) %>%
  summarize(nreads = prettyNum(
    sum(nreads) %/% if (seq_run[1] == "pb_500") 14 else 4,
    big.mark = " ")
  ) %>%
  deframe()
```

We attempted to denoise both long and short amplicons using `DADA2` version `r packageVersion("dada2")` [@callahan2016; @callahan2019] according to the steps outlined in  the supplementary information in @callahan2019.
However, only `r region_table$ASVs[region_table$region=="long"]` amplicon sequence variants (ASVs) were obtained for the long amplicons, representing `r round(100*region_table$reads[region_table$region == "long"]/as.integer(gsub(" ", "", demuxcounts["pb_500"])), 1)`% of the trimmed reads.
We believe that this poor performance was due to a combination of long read length and low sequencing depth relative to community diversity.
The `DADA2` algorithm requires that the seed sequence of each ASV be represented by at least two error-free reads [@callahan2016].
If sequencing errors are uniformly distributed, then the probability that a given read will be error-free is $(1-\epsilon)^L$, where $\epsilon$ is the sequencing error rate and $L$ is the read length in base pairs.
Then the number of reads of a given sequence that would be required to obtain two error-free reads in expectation is $2/(1-\epsilon)^L$.
For the combination of long reads (median $L=`r demuxlength['pb_500']` \text{ bp}$ after trimming) and moderate error rate (mean $\epsilon= `r format(demuxqual['pb_500'], digits = 4)`$ based on ccs quality scores) for the long amplicon in this study, the expected number of reads required to achieve two error-free reads is `r format(as.integer(round(2/(1 - demuxqual['pb_500'])^demuxlength['pb_500'])), big.mark = " ")`.
Given the high diversity relative to sequencing depth in this study (`r region_table$ASVs[region_table$seq_run == "pb_483" & region_table$region == "short"]` ASVs based on PacBio short amplicons, `r demuxcounts["pb_500"]` trimmed long amplicon reads), this requirement could not have been met for the long amplicons except by the most abundant sequences.
In comparison, the equivalent requirement for the short amplicon ($L=`r demuxlength['pb_483']` \text{ bp}$, $\epsilon=`r demuxqual['pb_483']`$) is only `r round(2/(1 - demuxqual['pb_483'])^demuxlength['pb_483'], 1)` reads.
We therefore developed a new workflow to assemble ASVs from the long amplicons, as follows:

@@

Taxonomic annotations of the Ribosomal Data Project's LSU fungal training set (RDP) version 11.5 [@cole2014] and Warcup ITS training set [@wang2007] were mapped to the taxonomic classification system used in the Unite database version 8 [@nilsson2019a].
In particular, the classification for fungi was according to @tedersoo2018, and for non-fungal eukaryotes was according to the proposed system of @tedersoo2017c.
Although the latter system is not formally published, it is consistent with the annotations for non-fungal eukaryotes in the Unite database.
Additionally, it is a system with both purportedly monophyletic taxa and a uniform set of taxon ranks, which make it more appropriate for sequence-based taxonomic assignment algorithms than more accepted classification systems such as that of the International Society of Protistologists [ @adl2019], which utilizes hierarchical nameless ranks.
FASTA format files of the re-annotated RDP and Warcup training sets are available at (*somewhere*).

@@

<!--An attempt was made to refine the long amplicon alignment using Pasta [@mirarab2014a], but the resulting maximum likelihood (ML) trees did not improve on the ML achieved using the original alignment from DECIPHER.
Since ML is the optimization criterion used by Pasta, the original DECIPHER alignment was retained.-->

@@

Long (short) amplicon sequencing with PacBio yielded `r rawcounts["pb_500"]` (`r rawcounts["pb_483"]`) circular consensus reads, of which `r demuxcounts["pb_500"]` (`r demuxcounts["pb_483"]`) were successfully demultiplexed.
The 5.8S region was detected in `r regioncounts["pb_500"]` (`r regioncounts["pb_483"]`) reads.
Ion Torrent sequencing yielded `r demuxcounts["is_057"]` demultiplexed reads, with the 5.8S region detected in `r regioncounts["is_057"]`.

(ref:readcounts) Number of Amplicon Sequence Variants (ASV) and reads at different pipeline stages.
Raw: raw reads as delivered by sequencing center;
Trim: reads with primers and demultiplexing indices removed;
LSUx: reads with a positive CM hit for 5.8S, allowing regions to be extracted;
ITS2: ASVs obtained for the ITS2 region, and reads successfully mapped to an ITS2 ASV;
short, ITS, LSU, long: ITS2-based ASVs mapped to consensus ASVs for longer regions, where \"short\" and \"song\" denote the full-length short and long amplicons, respectively.
Short amplicons were mapped to full ITS, LSU, and long amplicon ASVs when

```{r}
bioinf_table <- bind_rows(
  enframe(rawcounts, name = "seq_run", value = "reads") %>%
    mutate(
      reads = as.integer(gsub(" ", "", reads)),
      step = "Raw"
    ),
  enframe(demuxcounts, name = "seq_run", value = "reads") %>%
    mutate(
      reads = as.integer(gsub(" ", "", reads)),
      step = "Trim"
    ),
  enframe(regioncounts, name = "seq_run", value = "reads") %>%
    mutate(
      reads = as.integer(gsub(" ", "", reads)),
      step = "LSUx"
    ),
  filter(region_table, region == "ITS2") %>%
    select(seq_run, reads, step = region, ASVs),
  pivot_longer(asv_table, -1, names_to = "seq_run", values_to = "reads") %>%
    filter(reads > 0) %>%
    mutate_at("seq", chartr, old = "T", new = "U") %>%
    inner_join(
      select(readd(allseqs, cache = cache), seq = ITS2, long, short, ITS, LSU) %>%
        pivot_longer(-1, names_to = "region", values_to = "consensus") %>%
        filter(!is.na(consensus)) %>%
        select(-consensus) %>%
        unique(),
      by = "seq"
    ) %>%
    group_by(seq_run, region) %>%
    summarize(reads = sum(reads, na.rm = TRUE), ASVs = n()) %>%
    select(seq_run, reads, step = region, ASVs)
) %>%
  left_join(select(datasets, seq_run, tech, amplicon), by = "seq_run") %>%
  mutate(
    step = ifelse(tech == "Ion" & step == "Trim", "Raw", step),
  ) %>%
  select(tech, amplicon, step, reads, ASVs) %>%
  pivot_longer(c("reads", "ASVs"), names_to = "type", values_to = "count") %>%
  filter(!is.na(count)) %>%
  mutate(
    step = factor(step, levels = c("Raw", "Trim", "LSUx", "ITS2",
                                   "short", "ITS", "LSU", "long")),
    tech = factor(tech, levels = c("PacBio", "Ion", "Illumina")),
    amplicon = factor(amplicon, levels = c("Long", "Short")),
    type = factor(type, c("ASVs", "reads"))
  ) %>%
  arrange(tech, amplicon, type) %>%
  pivot_wider(
    id_cols = "step",
    names_from = c("tech", "amplicon", "type"),
    values_from = c("count")
    ) %>%
  arrange(step) %>%
  column_to_rownames("step")

options(knitr.kable.NA = '–')
bioinf_header <- tibble(name = names(bioinf_table)) %>%
  separate(name, into = c("tech", "amplicon", "type")) 
bioinf_table %>%
  set_names(gsub(".*_", "", names(.))) %>%
  kable(
    booktabs = TRUE,
    caption = "(ref:readcounts)",
    align = "rrrrrr",
    format.args = list(big.mark = "\\,"),
    linesep = c(rep("", 3), rep("\\addlinespace", 2), rep("", 4)),
    escape = FALSE
    ) %>%
  kableExtra::add_header_above(
    header = c(
      " " = 1,
      bioinf_header$amplicon[c(TRUE, FALSE)] %>% set_names(rep(2, length(.)), .)
    )
  ) %>%
  kableExtra::add_header_above(
    header = c(
      " " = 1,
      rle(bioinf_header$tech) %$% set_names(lengths, values)
    )
  )
```

@@

```{r pb-vs-it, fig.cap="(ref:readcap)", fig.height=8}
big_table <- bind_rows(
  mutate(asv_table, type = "ASV"),
  mutate(otu, type = "OTU")
) %>%
  select(-seq)

multi_table <- crossing(
  x_var = factor(datasets$seq_run, ordered = TRUE),
  y_var = factor(datasets$seq_run, ordered = TRUE),
  type = c("ASV", "OTU")
) %>%
  filter(
    x_var < y_var,
    x_var %in% names(big_table),
    y_var %in% names(big_table)
  ) %>%
  mutate_all(as.character) %>%
  group_by(x_var, y_var, type) %>%
  group_modify(
    function(d, g) {
      big_table %>%
        filter(type == g$type) %>%
        select(x = !!g$x_var, y = !!g$y_var) %>%
        filter(x > 0 | y > 0)
    }
  )

@@

```{r variofit}
variog_data %>%
  mutate_at("variofitST", lapply, purrr::compose(as.data.frame, as.list, gstat::extractPar)) %>%
  unnest(variofitST) %>%
  arrange(metric, amplicon, guild) %>%
  select(Guild = guild, Amplicon = amplicon, Metric = metric,
         Range = range, Anisotropy = anis) %>%
  mutate(Range = 3*Range,
         "Time Range" = Range/Anisotropy,
         "R/A" = ifelse(`R/A` > 100, NA, `R/A`)) %>%
  mutate_at(c("Range", "Time Range"), formatC, digits = 1,
            format = "f") %>%
  mutate_at("Range", paste, "m") %>%
  mutate_at("Anisotropy", paste, "m/a") %>%
  mutate_at("R/A", paste, "a") %>%
  mutate_at("R/A", str_replace, "NA a", "-") %>%
  mutate_at("Amplicon", fct_relabel, str_replace, " amplicon", "") %>%
  kable(booktabs = TRUE, caption = "(ref:variofit)")
```

@@

```{r phylotax, fig.height = 2, fig.width = 3, fig.cap = "(ref:phylotax)"}
extree <-
  treeio::as.treedata(ape::read.tree(text = "(A:1,((B:1,C:1):1,((E:1,F:1):1,D:1):1):1);")) %>%
  tidytree::full_join(
    tibble(label = LETTERS[1:6],
           tax1 = c("unk", "Tax1", "Tax2", "Tax2", "unk", "unk"),
           tax2 = c("unk", "Tax2", "Tax2", "Tax1", "unk", "Tax1")),
    by = "label"
  )
ggtree::ggtree(extree) +
  ggtree::geom_tiplab(x = 4, align = 2) +
  xlim(0, 6) +
  ggtree::geom_tiplab(
    aes(color = tax1, label = tax1),
    x = 4.7,
    hjust = 0.5
  ) +
  ggtree::geom_tiplab(
    aes(color = tax2, label = tax2),
    x = 5.7,
    hjust = 0.5
  ) +
  ggtree::geom_tiplab(x = 5.2, label = "/", hjust = 0.5) +
  ggtree::geom_nodelab(aes(label = node - 6), geom = "label") +
  ggtree::geom_hilight(node = 10, fill = "steelblue", extendto = 4) +
  ggtree::geom_hilight(node = 9, fill = "tomato", extendto = 4) +
  scale_color_manual(values = c(Tax1 = "steelblue", Tax2 = "tomato", unk = "grey50")) +
  scale_y_reverse()
```
